---
title: "RTMLE for dummies"
author: "Christian Torp-Pedersen"
date: "`r Sys.Date()`"
output: pdf_document
---

RTMLE is a package for conducting LTMLE analyses with a range of modifications targeted for register based data. Many features of the package are developed for ease of use in comparison with the LTMLE package and no claim of superiority is made. The scope is to enable trial emulations based on observational data. Target trial emulations with observational data are complicated by the fact that interventions in clincial practice differ from randomised trials. Often adherence to an intervention can vary substantially and therefore relying on "starting treatment" as a proxy for "intention to treat" can result in misleading results, in particular when adherence is very low.  

Learning and understanding basics of LTMLE analyses is best obtained from "Targeted Learning in Data Science: Causal Inference for Complex Longitudinal Studies", Springer Series in Statistics), van der Laan & Rose, 1st ed. 2018 Edition.  There are numerous papers discussing selected strategies related to the technique.

The following is a simplistic presentation of LTMLE/RTMLE for non statistician users that intend to emulate a trial.

A longitudinal study is considered where participants at time zero receive one of two treatments. Subject selection may rely on starting a treatment of interest versus a selected control treatment or it may rely on comparing subjects starting a selected treatment with controls selected by a variety of mechanisms which could be matching.  Note that if matching users to non-users what is calculated is average treatment effect of the treated rather than average treatment effect.

The follow-up time of interest is divided in a series of time slices of equal length. The length of these time slices need to be short enough to ensure biological meaning and long enough to ensure that all combinations of variables in each time slice can be found with a probability above zero and below one (positivity assumption). If the exposure of interest relies on prescribed drugs, then the time slices should reflect meaningful periods seen in clinical practice such as a single month up to a year or more. 

During each time slice a subject may be censored (observation time ends, subject disappears), have an outcome of interest or be subject to a competing risk (typically death unrelated to outcome).

For each time slice all variables of interest needs to be summarized to a category or value. Intermittent treatment needs to be summarized as "treatment" or "no treatment", varying values summarized to a single value.

In addition to specifying variables during periods the user needs to specify which estimand to calculate.  With a single treatment of interest a typical estimand could be "Always treat with A" versus "never treat with A". If there is a control treatment the estimand can be more complex such as: "Always treat with A and never with B" versus "Always treat with B and never with A".  During follow-up it may be of interest to specify only treatment during the first few periods or treatment consistently during follow-up.  Such possibilities have been implemented.  This can be important if outcomes later than the treatment period are of interest.  The LTMLE method also allows continued treatment to rely on probabilities which could be used to approximate real user discontinuation, but such features have not yet been practically implemented.

The package does regression which takes into account both the exposure/covariate relation and the exposure outcome relation, an ability named "double robust". Thus for each time slice the probability of exposure during the next time slice is calculated and outcome during the next period can be determined from exposure variables from all preceding periods. In reality the regression starts from the last periods and moves backward using a regression technique developed by Robins (ref).

A strength of the method is that the regressions to obtain propensity of treatment as well as outcome can be very flexible and include a library of regression methods.  For each project a "superlearner" can be specified to include parametric as well as non-parametric methods.  Choice of regression methods can be complex, but it is wise to include penalized regression parametric methods in order to downplay non-essential variables and also wise to include a tree based methods such as random forest in order to capture critical interactions with treatments.  The final comparison is from G-estimation simulating a situation where all individuals first gets the first treatment choice and thereafter the other treatment. In this process all other variables than treatment are inherited and therefore interaction with treatment cannot be realized with specifying interaction variables in models.

What the method provides eventually is a hypothetical randomized study where the whole population (both treatment groups) first go through follow-up on one intervention and afterwards start over with the alternative.  This is termed G-estimation.

LTMLE is a method in the domain of "causal inference" and needs to rely on the basic assumptions of such methods: Exchangeability, consistency and positivity.  The exchangeability assumption specifies that switching treatment should provide a mirror result and is also phrased as an assumption of having all relevant confounders properly specified.  The consistency assumption specifies that the effect of intervention is independent on how is was provided. This may be obvious for a medical therapy with tablets but can be complex in other situations. The final assumption is positivity which has already been mentioned and which specifies that all variable combinations in each time slice should have probabilities greater than zero and less than one.  

If there are many variables and many time slices it is easy to end with a situation where positivity violation results in extreme confidence limits or crashing models.  Two remedies needs consideration.  First, variable selection should be careful.  If a condition is associated with for example multiple varying treatments specifed as independent variables, then it is almost inevitable that positivity violations will occur during some time slices.  The second option and the one recommended by those developing LTMLE is simulations.  A range of simulation studies not using the actual outcome are conducted and only when models run smoothly with simulated outcomes is the final model examined.

If the variables considerations are not followed it is easy to end in an unacceptable situation where variables are added and removed until a sensible model is found.  Such an approach is highly biased and should be avoided.  The proper approach is to specify the path in an analysis plan and then also specify eventually those steps where the analysis plan needed to be modified during calculations.

# Data preparation
The final input to the rtmle function is quite complex and provides time varying data
in a wide format as lists. The method described here to obtain properly formatted data is
only one method possible. The method relies on keeping data in a long format as far as possible to make it feasible to check that each step is done correctly

## Step 1 - Get exercise data

## Baseline data and outcomes

The example has a single baseline variable (Age), Outcome_time, Trial_Start and
Event_Type (0=censor, 1=event, 2=competing risk).

```{r, echo=FALSE}
library(data.table)
library(rtmle)
# set working directory 
data(baseline_data)
data(treatment_data)
data(outcome_data) 
head(baseline_data)
```

The next dataset has treatment periods where patients are allocated to be treated with 
"A" or "B" - some consistently, others discontinue

```{r, echo=FALSE}
head(treatment_data)
```
The final dataset are the outcome data. Note that variables for event of interest,
censoring and competing risk are coded with separate variables. For each outcome these three variables are mutually exclusive.  If there are multiple outcomes it can be recommended to handle 
them in a single dataset for data management.
```{r, echo=FALSE}
head(outcome_data)
```


# Prepare for LTMLE
LTMLE eventually requires that information is provided in a wide format with one record per individual and variables for each covariate.  The following procedure allows most of the management to be conducted in a long formate which eases checking of programming.

The basis is the number of time slices and in this example there are four, of which two are used in calculations.

During the data preparation below, records are split into multiple records and 
to avoid confusion with the original entry into analysis, a new variables (inn/out)
are defined and used in further data management.

We start with "base" which holdes the ID, the start and the end. Because of the splitting we copy Start/End to inn/out for splitting purposes.  Note that all participants are required to have information for all periods even when they stop early in the study - information after stopping will not be used in calculations.
```{r, echo=TRUE}
base <- baseline_data[,.(ID,Trial_Start)]
base[,':='(inn=Trial_Start,out=Trial_Start+4*365)] # Four periods of 365 days
```
## Splitting
The following steps have the purpose of defining levels of variables in each of 
the defined time periods (tine slices). To start this process all records are split according 
to timing of change in variable status. The order of splitting is not important.

The first step is splitting by all variables that change only once. 
This is performed with the splitTwo function that needs the original 
base data and a "splitting guide" which is the dataset with dates where variables
that change only once are held.  For the current example the only variables are censoring, outcome and competing risk nodes.  These variables are therefore defined in distinct variables. The list of variables can for realistic examples also include other time dependent variables.  Note that the content of each variable that is used for splitting needs to have value (numeric or date) at the time of change, otherwise NA.

Note: If there are multiple outcomes for study these can be be handled simultaneously with organised naming.

```{r echo=TRUE}

longSplit <- splitTwo(indat=base,
                      splitdat=outcome_data,
                      invars=c('ID','inn','out'),
                      splitvars = c('ID','censor','outcome','compete'))
```

Next the data is split by the time dependent variables with potentially multiple
changes during the study. We use the function heaven::splitFromTo. A single call to the function can split on all time dependent variables representing intervals. Apart from variables indicating start and end of periods, two more variables are needed. One variable indicates a name for the condition (here "treatment") and the other a "value" for that treatment period. The function does not allow overlap whithin person/condition. This needs to be arranged prior to use of the function.

```{r echo=TRUE}
splitguideTime <- copy(treatment_data)
splitguideTime[,value:=1]
longSplit <- splitFromTo(indat=longSplit,
                         splitdat=splitguideTime,
                         invars=c('ID','inn','out'),
                         splitvars =      c('ID','Treatment_Start','Treatment_End','value','treatment'))

```

Finally, the data is split by the selected time periods, in the current example
just five periods. The new value 'period' contains the period number. This uses
heaven::splitSeq.

```{r echo=TRUE}
longSplit <- splitSeq(indat=longSplit,
                      invars=c('ID','inn','out'),
                      varname = 'Trial_Start', # intervals since inclusion in study
                      splitvector= seq(-1,5*365-1,365), # just five periods for the example
                      format = "vector",
                      value="period")
```
## Summarize in periods

With the splitting complete, all information for each selected equally
sized time period (in the example four) is separated.

The next step is then to summarize information by 'period'.

The outcomes (event, censoring, competing risk) should be the maximal
outcome for each period since an outcome event is coded "1" as opposed
to "0" when not occurring.

For other variables the chosen summary should reflect relevant biology. It could be
the value at exit, at entry, rely on percentage of exposure during the period etc.
For the current simplistic example we will use any exposure during a period as a predictor
for the next period.

```{r echo=TRUE}
setkeyv(longSplit,c("ID","period","inn"))
# Max value of outcomes and value of time dependent variables at period entry
longSplit <-longSplit[,':='(outcome=max(outcome),censor=max(censor),compete=max(compete),
                A=A[1],B=B[1]),
          by=c("ID","period")] 
setkeyv(longSplit,c("ID","inn"))
# Choose first record for each ID/period
aggrSplit <- longSplit[,.SD[1],by=c("ID","period")]
aggrSplit[,':='(A=as.numeric(A),B=as.numeric(B))]

# Exposure needs to come before outcome, so the exposure is moved on period back. 
# This results in the exposure during the first time period to become baseline exposure
# which may be correct in one situation and wrong in others. The example is chosen to make
# this correct.
aggrSplit_cov <- aggrSplit[,.(ID,period,A,B)] # all time dependent variables, here just "treatment"
aggrSplit_cov[,period:=period-1]
aggrSplit_out <- aggrSplit[,.(ID,period,outcome,censor,compete)] # outcomes
```
## Transpose to wide format
```{r echo=TRUE}
outcome_dt <- dcast(aggrSplit_out,ID~period,value.var = c("compete","censor","outcome"))
# For the time dependent covariates we need a list with one member for each time dependent covariate
treatment_dt <- longToWideList(aggrSplit_cov,"ID",c("A","B"))
```


## LTMLE - one variable for treatment, here "A". The comparison is for the target parameter "Always treat with A versus never treat with A".

```{r echo=TRUE}

x <- rtmle_init(intervals=4,name_id='ID',name_time='period',name_outcome='outcome',
                name_competing='compete',name_censoring='censor',
                censored_levels=c('1','0'),censored_label="1")
x$data <- list(baseline_data = baseline_data[,.(ID,age)],
               outcome_data = outcome_dt,
               timevar_data=treatment_dt[1] # just "A"
               )
prepare_data(x) <- list()
protocol(x) <- list(name = "A",treatment_variables = "A",intervention = 1)
protocol(x) <- list(name = "not A",treatment_variables = "A",intervention = 0)
target(x) <- list(name = "Risk",strategy = "additive",estimator = "tmle",protocols = c("A","not A"))
x <- run_rtmle(x,time_horizon = 2)
summary(x)

```
## LTMLE - Same analysis, but this time a complex superlearner with glmnet and to versions of random forest.
```{r echo=TRUE}
x <- rtmle_init(intervals=4,name_id='ID',name_time='period',name_outcome='outcome',
                name_competing='compete',name_censoring='censor',
                censored_levels=c('1','0'),censored_label="1")
x$data <- list(baseline_data = baseline_data[,.(ID,age)],
               outcome_data = outcome_dt,
               timevar_data=treatment_dt[1] # just "A"
)
prepare_data(x) <- list()
protocol(x) <- list(name = "A",treatment_variables = "A",intervention = 1)
protocol(x) <- list(name = "not A",treatment_variables = "A",intervention = 0)
target(x) <- list(name = "Risk",strategy = "additive",estimator = "tmle",protocols = c("A","not A"))
x <- run_rtmle(x,
               time_horizon = 2,
               refit = TRUE,
               learner = list("learn_ranger_50" = list(num.trees = 20,learner_fun = "learn_ranger"),
                              "learn_glmnet"),folds = 10)
summary(x)
```


## LTMLE - Two variable for treatment, A and B. The comparison is for the target parameter "Always treat with A and never B versus Always treat with B and never with A".

```{r echo=TRUE}

x <- rtmle_init(intervals=4,name_id='ID',name_time='period',name_outcome='outcome',
                name_competing='compete',name_censoring='censor',
                censored_levels=c('1','0'),censored_label="1")
x$data <- list(baseline_data = baseline_data[,.(ID,age)],
               outcome_data = outcome_dt,
               timevar_data=treatment_dt
               )
prepare_data(x) <- list()

protocol(x) <- list(name = "Always_A_never_B",
                    intervention = data.frame("A" = factor("1",levels = c("0","1")),
                                              "B" = factor("0",levels = c("0","1"))))
protocol(x) <- list(name = "Always_B_never_A",
                    intervention = data.frame("A" = factor("0",levels = c("0","1")),
                                              "B" = factor("1",levels = c("0","1"))))
x$protocols
target(x) <- list(name = "Risk",strategy = "additive",estimator = "tmle",protocols = c("Always_A_never_B","Always_B_never_A"))
x <- run_rtmle(x,time_horizon = 2)
summary(x)

```
## LTMLE - Two variable for treatment, A and B. The comparison is for the target parameter "Always treat with A and never B versus Always treat with B and never with A".

BUT, this time the treatment is only compared for period one with a requirement of no
treatment in period 2

```{r echo=TRUE}

x <- rtmle_init(intervals=4,name_id='ID',name_time='period',name_outcome='outcome',
                name_competing='compete',name_censoring='censor',
                censored_levels=c('1','0'),censored_label="1")
x$data <- list(baseline_data = baseline_data[,.(ID,age)],
               outcome_data = outcome_dt,
               timevar_data=treatment_dt
               )
prepare_data(x) <- list()

protocol(x) <- list(name = "Always_A_never_B",
                    intervention = data.frame("A" = factor(c("1","0"),levels = c("0","1")),
                                              "B" = factor(c("0","0"),levels = c("0","1"))))
protocol(x) <- list(name = "Always_B_never_A",
                    intervention = data.frame("A" = factor(c("0","0"),levels = c("0","1")),
                                              "B" = factor(c("1","0"),levels = c("0","1"))))
x$protocols
target(x) <- list(name = "Risk",strategy = "additive",estimator = "tmle",protocols = c("Always_A_never_B","Always_B_never_A"))
x <- run_rtmle(x,time_horizon = 2)
summary(x)

```
